---
title: "Assignment 4: Exploratory Data Analysis"
author: "Kent Maramba"
date: "`09/25/2026`"
format: html
editor: visual
---

## Overview

Due: **Friday 9/25 11:59pm**

In this assignment, you will practice Exploratory Data Analysis (EDA) by making plots from a **Kaggle dataset of your choice**.

You will submit **at least 4 different plot types**, with short written interpretations for each.

Part 1 is worth a total of 1 point, and Part 2 is worth 4 points (1 point for each plot, question, and interpretation).

## Part 1: Get Your Data

1.  Go to [Kaggle Datasets](https://www.kaggle.com/datasets), create an account, and choose a dataset that interests you. Ensure your data has:
    1.  At least 4 variables
    2.  At least one categorical and one continuous variable
    3.  I recommend data that contains at least 50-100 observations
    4.  Tip: Open the CSV on Kaggle first and skim the columns to check that it meets these requirements
2.  Download the dataset as a CSV file.
3.  Place the CSV in your class R project `data` folder.
4.  Read it into R with `read_csv()`.

```{r, message=FALSE}
library(tidyverse)
mydata <- read_csv("data/tips.csv")

# mydata <- read_csv("data/your_dataset.csv")

```

### Describe your data

Data author: Saurabh Badole

Link: <https://www.kaggle.com/datasets/saurabhbadole/restaurant-tips-dataset/data>

Description: This data set contains information about restaurant transactions, including total bills, tips, customer gender, smoking preference, dining day, time of day, and party size. It can be used to analyze tipping behavior and identify factors that may influence the amount of tips customers leave.

### Preview your data

Use some of the strategies you've learned in this class or picked up elsewhere to summarize and quickly explore your data.

```{r}
summary(mydata)
```

### Summary stats

Before you make plots, you must also:

- Show basic summaries of your dataset, including:

  - `count()` for categorical variables (to see category frequencies)

  - `mean()`, `median()`, `sd()`, `min()`, `max()` for at least one numeric variable

  - Remember to use `na.rm = TRUE` if you have missing values that affect summaries!

```{r}
head(mydata)
count(mydata, sex)

mean(mydata$total_bill)
median(mydata$total_bill)
sd(mydata$total_bill)
min(mydata$total_bill)
max(mydata$total_bill)
```

Write 2–3 sentences here: What do the category counts tell you? What does the numeric summary show about spread, center, or outliers?

Based on the category count, there are more males than female customers recorded paying the bill. The average bill was about \$19.79 while the median was \$17.80 with the bills ranging from \$3.07-\$50.81. Standard deviation is 8.90 meaning the bills had noticeable amount of variation.

Part 2: Visualize Your Data

Create **at least 4 plots**, each of a different type. Your plots may share variables, but each one should answer a **different question** about your data. At least one plot must include a variable that isn't in the others.

Each plot should have:

- A title and labeled axes

- A short interpretation (2-3 sentences) underneath

  - Sample interpretation: (boxplot of `price` by `cut` from `diamonds` data): *Ideal diamonds have the lowest median price, which is surprising since Ideal is the best cut. Every cut also has a lot of very expensive outliers. One possible reason for the low Ideal prices is that Ideal diamonds might be smaller on average, which I could check by plotting carat by cut.*

Use the EDA Part 3 notes as your reference. Since this is an EDA assignment, your plots can look plain, but consider adding colors, themes, or interactivity for interpretability and visual appeal if you know how.

### Plot 1

```{r}
# insert code here
ggplot(mydata, aes(x=day))+
  geom_bar()
```

Question this plot answers: Which days had the most records of bills?

Interpretation: From this bar graph, I see that Saturday, Sunday(weekends) and Thursdays had the most bills. Fri being the lowest recorded. Rest of the weekdays had no records of service.

### Plot 2

```{r}
# insert code here
ggplot(mydata, aes(x= total_bill, y = tip))+
  geom_point()
```

Question this plot answers: Do people with high bill totals leave higher tips?

Interpretation: the scatter plot shows that as the bill increases, generally the tip does too increase but the points are spread out. Also, there are a lot of bills with less that \$30.

### Plot 3

```{r}
# insert code here
ggplot(mydata, aes(x=total_bill,))+
  geom_histogram()+
  labs(title = "Distribution of Total Bills", x= "total bills($)", y= "Amount of Bills")
```

Question this plot answers: What range of total bill amounts is most common?

Interpretation: Majority of the bills recorded fall in the \$10-\$25 range.

### Plot 4

```{r}
# insert code here
ggplot(mydata, aes(x=smoker, y=tip))+
  geom_boxplot()
```

Question this plot answers: Do smokers and non-smokers tip differently?

Interpretation:

Smokers and non-smokers generally had similar tips with the smokers having a slight upperhand with comes to high tips. Because its so close I would not claim they do because of the similarity. However both groups also have unusual high tips as outliers.

------------------------------------------------------------------------

# Push to GitHub AND submit link on Canvas!!

Assignment is due Friday 9/25 at 11:59 pm.
