---
title: "CS-201 First Look at R"
author: "Kent Maramba"
date: 8/26/26
format: html
editor: visual
toc: true
---

This is the first class session walking through code together!

We will be programming in R and in this class will primarily use `qmd` documents to keep things tidy and presentable. Each code chunk below is where you will run R code. Feel free to add notes as we go along.

## Navigating RStudio

By default, everyone should have the same four-pane layout:

1.  **Top left**: Source editor

2.  **Bottom left**: Console (and Terminal and Background Jobs)

3.  **Top Right**: Environment

4.  **Bottom Right**: Files, Plots, Help, etc.

### Customizing look & feel

Feel free to customize your RStudio.\
Go to Tools \> Global Options \> Appearance to select a color scheme you like.

I also personally like to enable these:

- Tools \> Global Options \> Code \> Insert matching parens/quotes

- Tools \> Global Options \> Console \> Show syntax highlighting in console input

I don't recommend rearranging your pane layout at this time.

## Running Code

Try running some code

Load Library

```{r}
library(ggplot2)

```

Some Scripting

```{r}
# this is a comment 
quiz1 <- 88 #this is how to assign a value to a name, also command option i
quiz2 <- 94
quiz3 <- 79

quiz1

#average
(quiz1 + quiz2 + quiz3) / 3

mean(quiz1, quiz2, quiz3) #not correct justs add everything up
mean(quiz1 + quiz2 + quiz3) #not correct only returns 1st argument

mean(c(quiz1,quiz2, quiz3))
students <-c("a", "b", "c")
list <- c(quiz1, quiz2, quiz3)
scores <-c(88, 94, 79)
barplot(scores, names.arg = students)

barplot(scores, names.arg = students, ylim = c(0, 100), main = "TITLE", xlab = "pupils", y = "what they got")
```

\

## Introducing `ggplot2`

#### Make a data frame

```{r}
grades_df <- data.frame(student = students, score = scores)

grades_df

summary(grades_df)
dim(grades_df) # ndf of dimensons 
ncol(grades_df)# number of colums
nrow(grades_df) #number of rows
```

Try ggplot

```{r}
ggplot(data = grades_df, aes(x = student, y = score)) #just the foundation or 1st layer

ggplot(data = grades_df, aes(x = student, y = score, group = 1)) + 
  geom_col(fill= "steelblue") + 
  geom_point() +
  geom_line() +
  theme_minimal()+
  labs(title = "BEST GRAPH EWA",
       subtitle = "some quiz scores",
       x  = "da best students",
       y= "what they scored on the quiz")
       
       


```

\

mutate the mainframe

```{r}
library(tidyverse)

grades_df <- grades_df |>
  mutate(study_hours = c(2, 5, 1))

```

Make a scatter plot

```{r}
ggplot(data = grades_df, aes(x = study_hours, y = score)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  geom_blank()
```

## Extra Credit (1pt)

What result do you expect when you run this expression?

```{r}
1+1+1
quiz1 + quiz2
1+1+1-3

0.1 + 0.1 + 0.1 -0.3

options(scipen= 999)
print (0.1, 20)
print(0.2, 20)
print(0.3,20)
```

What did you observe?

> Answer:

Why do you think this is happening? (You can look online for help)

> Answer:
